---
title: Pedigree alignment details
author: TM Therneau
date: '`r format(Sys.time(),"%d %B, %Y")`'
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
header-includes: \usepackage{tabularx}
vignette: |
  %\VignetteIndexEntry{Pedigree alignment details}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, library_charge}
library(Pedixplorer)
```
    
Pedigree alignment
==========================

An **aligned** Pedigree is an object that contains a Pedigree along
with a set of information that allows for pretty plotting.
This information consists of two parts: 
a set of vertical and horizontal plotting coordinates along with the
identifier of the subject to be plotted at each position,
and a list of connections to be made between parent/child, spouse/spouse,
and twin/twin.

Creating this aligment turned out to be one of the more difficult parts
of the project, and is the area where significant further work could be
done.  

All the routines in this section completely ignore the $id$ component
of a Pedigree; everyone is indexed solely by their row number in the object.

Hints
==========================

The first part of the work has to do with a **hints** list for each
Pedigree.  It consists of 3 parts:

- The left to right order in which founders should be processed.
- The order in which siblings should be listed within a family.
- For selected spouse pairs, who is on the left/right, and which of the
two should be the anchor, i.e., determine where the marriage is plotted.

The default starting values for all of these are simple: founders are 
processed in the order in which they appear in the data set, 
children appear in the order they are found in the data set,
husbands are to the left of their wives, and a marriage is plotted
at the leftmost spouse.
A simple example where we want to bend these rules is when two families
marry, and the Pedigrees for both extend above the wedded pair.  
In the joint Pedigree the
pair should appear as the right-most child in the left hand family, and
as the left-most child in the right hand family.
With respect to founders, assume that a family has three lineages with
a marriage between 1 and 2, and another between 2 and 3.  In the joint
Pedigree the sets should be 1, 2, 3 from left to right.  

The hints consist of a list with two components.

1. A vector of numbers of the same length as the Pedigree,
used to order the female founders and to order siblings within
family.  For subjects not part of either of these the value can be 
arbitrary.  
2. A 3 column matrix of spouse pairs, each row indicates the
left-hand member of the pair, the right-hand member, and which of the two
is the anchor, i.e., directly connected to their parent.  
Double and triple marriages can start to get interesting.

Auto_hint
==========================

This routine is used to create an initial hints list. It is a part of
the general intention to make the routine do pretty good drawings
automatically.
The basic algorithm is trial and error. 

- Start with the simplest possible hints (user input is accepted)
- Call align to see how this works out
- Fix any spouses that are not next to each other but could be.
- Any fix on the top level mixes up everything below, so we do the
fixes one level at a time.

The routine makes no attempt to reorder founders. It just is not smart enough
to figure that out.

The first thing to be done is to check on twins. They increase the complexity,
since twins need to move together.
The `rel(ped, "code")` object is a factor, so first turn that into numeric.
We create 3 vectors:

- $twinrel$ is a matrix containing pairs of twins and
their relation, it is a subset of the incoming $rel$ dataframe.
- $twinset$ vector identifies twins, it is 0 for anyone who is not a 
part of a multiple-birth set, and a unique id for each member of a set.  
We use the minimum row number of the members of the set as the id.
- $twinord$ is a starting order vector for the set; it mostly makes sure
that there are no ties (who knows what a user may have used for starting 
values.)  

A recent addition is to carry forward packaged and align to 
kindepth and align.

## Shift in auto_hint

Next is an internal function that  rearranges someone to be
the leftmost or rightmost of his/her siblings.  The only
real complication is twins. If one of them moves the other has to move too.  
And we need to keep the monozygotics together within a band of triplets.

*Algorithm* : if the person to be moved is part of a twinset, 
first move all the twins to the left end (or right
as the case may be), then move all the monozygotes to the
left, then move the subject himself to the left.

## Init auto_hint

Now, get an ordering of the Pedigree to use as the starting point.  
The numbers start at 1 on each level. We do not need the final *prettify* step,
hence `align = FALSE`. If there is a hints structure entered, we retain its
non-zero entries, otherwise people are put into the order of the data set.

We allow the hints input to be only an order vector. twins are then further
reordered.

## Fixup auto_hint

The result coming back from `align()` is a set of vectors and
matrices:

- $n$ vector, number of entries per level
- $nid$ matrix, one row per level, numeric id of the subject plotted here
- $spouse$ integer matrix, one row per level, subject directly to my
right is my spouse (1), a double marriage (2), or neither (0).
- $fam$ matrix, link upward to my parents, or 0 if no link.

```{r, auto_hint1, echo = FALSE, fig.dim = c(10, 8)}
#
# A simple Pedigree to illustrate autohint's code
#
test1 <- data.frame(id = 1:11,
    sex = c("m", "f")[c(1, 2, 1, 2, 1, 1, 2, 2, 1, 2, 1)],
    dadid = c(0, 0, 1, 1, 1, 0, 0, 6, 6, 6, 9),
    momid = c(0, 0, 2, 2, 2, 0, 0, 7, 7, 7, 4)
)
ped1 <- Pedigree(test1, missid = "0")

temp2 <- Pedigree(test1, missid = "0", hints = list(
    horder = setNames(
        c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11),
        1:11
    )
))

par(mfrow = c(1, 2))
plot(temp2, title = "Before auto_hint")
plot(ped1, title = "After auto_hint")
```

Now, walk down through the levels one by one.
A candidate subject is one who appears twice on the level, once
under his/her parents and once somewhere else as a spouse.
Move this person and spouse the the ends of their sibships and
add a marriage hint.
The figure above shows a simple case.  The input data set has
the subjects ordered from 1--11, the left panel is the result without
hints which processes subjects in the order encountered.
The return values from [align] have subject 9 shown twice.
The first is when he is recognized as the spouse of subject 4, the second
as the child of 6--7.

The basic logic is

- Find a subject listed multiple times on a line (assume it is a male).
    This means that he has multiple connections, usually one to his parents and
    the other to a spouse tied to her parents.  (If the
    spouse were a marry-in she would have been placed alongside and there
    would be no duplication.)
- Say subject x is listed at locations 2, 8, and 12.  We look at one
    pairing at a time, either 2-8 or 8-12.  Consider the first one.
    - If position 2 is associated with siblings, rearrange them to
        put subject 2 on the right.  If it is associated with a spouse at
        this location, put that spouse on the right of her siblings.
    - Repeat the work for position 8, but moving targets to the left.
    - At either position, if it is associated with a spouse then
        add a marriage.  If both ends of the marriage are anchored, i.e.,
        connected to a family, then either end may be listed as the anchor
        in the output; follow the suggestion of the duporder routine.  If
        only one is, it is usually better to anchor it there, so that the
        marriage is processed by [align] when that family is.
        (At least I think so.)

This logic works 9 times out of 10, at least for human pedigrees.
We'll look at more complex cases below when looking at the $duporder$ (order
the duplicates) function, which returns a matrix with columns 1 and 2 being a
pair of duplicates, and 3 a direction. Note that in the following code
$idlist$ refers to the row numbers of each subject in the Pedigree, not to
their label `ped(ped, "id")`.

## duporder

For the case shown in figure above the [duporder] function
will return a single row array with values `(2, 6, 1)`, the first two
being the positions of the duplicated subject.  
The anchor will be 2 since that is the copy connected to parents
The direction is TRUE, since the spouse is to the left of the anchor point.
The id is 9, sibs are 8, 9, 10, and the shift function will create position
hints of 2,1,3, which will cause them to be listed in the order 9, 8, 10.

The value of spouse is 3 (third position in the row), subjects 3, 4, and 5
are reordered, and finally the line `(4,9,1)` is added to the sptemp 
matrix.  
In this particular case the final element could be a 1 or a 2, since both
are connected to their parents.

```{r, align2, echo = FALSE, fig.dim = c(10, 8)}
#
# The second, more complex test Pedigree
#
test2 <- data.frame(id = c(1:13, 21:41),
    dadid = c(0, 0, 1, 1, 1, 0, 0, 6, 6, 6, 0, 11, 11,
        0, 0, 0, 0, 0, 0, 21, 21, 21, 23, 23, 25, 28,
        28, 28, 28, 32, 32, 32, 32, 33
    ),
    momid = c(0, 0, 2, 2, 2, 0, 0, 7, 7, 7, 0, 5, 9,
        0, 0, 0, 0, 0, 0, 22, 22, 22, 24, 24, 26, 31,
        31, 31, 31, 29, 29, 29, 29, 13
    ),
    sex = c(1, 2, 1, 2, 2, 1, 2, 1, 2, 1, 1, 1, 2, 1,
        2, 1, 2, 1, 2, 1, 1, 2, 2, 2, 1, 1, 1, 2, 2,
        1, 2, 1, 1, 2
    )
)
ped2 <- Pedigree(test2, missid = "0")
ped2a <- Pedigree(test2, missid = "0", hints = list(
    horder = setNames(seq_along(test2$id), test2$id)
))
par(mfrow = c(1, 2))
plot(ped2a, title = "Before auto_hint")
plot(ped2, title = "After auto_hint")
```

The figure above shows a more complex case with several arcs.
In the upper left is a double marry-in.
The $anchor$ variable in the above code
will be `(2,2)` since both copies have an anchored spouse.
The left and right sets of sibs are reordered (even though the left
one does not need it), and two lines are added to the sptemp matrix:
`(5,11,1)` and `(11,9,2)`.

On the upper right is a pair of overlapping arcs.
In the final tree we want to put sibling 28 to the right of 29 since
that will allow one node to join, but if we process the subjects in
lexical order the code will first shift 28 to the right and then later
shift over 29.
The duporder function tries to order the duplicates into a matrix
so that the closest ones are processed last.  The definition of close
is based first on whether the families touch, and second on the
actual distance.
The third column of the matrix hints at whether the marriage should
be plotted at the left (1) or right (2) position of the pair.  The
goal for this is to spread apart families of cousins; in the
example to not have the children of 28/31 plotted under the 21/22
grandparents, and those for 29/32 under the 25/26 grandparents. 
The logic for this column is very ad hoc: put children near the edges.

## Find-spouse and find-sibs

Finally, here are two helper routines.
Finding my spouse can be interesting -- suppose we have a listing with
Shirley, Fred, Carl, me on the line with the first three marked as
`spouse = TRUE` -- it means that she has been married to all 3 of us.
First we find the string from rpos to lpos that is a marriage block;
99\% of the time this will be of length 2 of course.  Then find
the person in that block who is opposite sex, and check that they
are connected.
The routine is called with a left-right position in the alignment
arrays and returns a position.

The findsibs function starts with a position and returns a position as well,
and is much simpler than findspouse.

## Fixup2

At this point the most common situation will be what is shown in 
figure.  The variable $anchor$ is `(2,1)` showing that the
left hand copy of subject 9 is connected to an anchored spouse and the
right hand copy is himself anchored.  The proper addition to the
spouselist is `(4, 9, dpairs)`, where the last is the hint from the
dpairs routine as to which of the parents is the one to follow further when
drawing the entire Pedigree.  (When drawing a Pedigree and there is a
child who can be reached from multiple founders, we only want to find
the child once.) 

The double marry-in found in the figure, subject 11, leads
to value of `(2,2)` for the [anchor] variable.  The proper addition to
the $sptemp$ matrix in this case will be two rows, `(5, 11, 1)` indicating
that 5 should be plotted left of 11 for the 5-11 marriage, with the first
partner as the anchor, and a second row `(11, 9, 2)`.
This will cause the common spouse to be plotted in the middle.

```{r, align3, echo = FALSE, fig.dim = c(10, 8)}
#
# A simple pedigree to illustrate autohint's code
#
test3 <- data.frame(id = 1:14,
    sex = c("m", "f")[c(2, 1, 1, 1, 1, 1, 2,
        2, 1, 2, 1, 2, 2, 1
    )],
    dadid = c(0, 6, 2, 2, 2, 0, 0, 11, 0, 9, 6, 0, 6, 11),
    momid = c(0, 7, 1, 1, 1, 0, 0, 12, 0, 1, 7, 0, 7, 12),
    affected = c(0, 1, 1, 0, 0, 0, 0, 0, 0, 0, 1, 0, 0, 0)
)
ped3 <- Pedigree(test3, missid = "0", hints = list(
    horder = setNames(1:14, 1:14)
))

test4 <- data.frame(id = 1:17,
    sex = c("m", "f")[c(2, 1, 1, 2, 2, 1, 2, 1,
        2, 2, 1, 2, 2, 1, 1, 2, 2
    )],
    dadid = c(0, 8, 6, 6, 0, 0, 0, 0, 11,
        0, 0, 6, 0, 0, 0, 11, 14
    ),
    momid = c(0, 9, 5, 5, 0, 0, 0, 0, 10,
        0, 0, 7, 0, 0, 0, 12, 5
    )
)

ped4 <- Pedigree(test4, missid = "0", hints = list(
    horder = setNames(1:17, 1:17)
))

par(mfrow = c(1, 2))
plot(ped4)
plot(ped3)
```

Multiple marriages can lead to unanchored subjects.  
In the left hand portion of the figure above we have two
double marriages, one on the left and one on the right with 
anchor values of `(0,2)` and `(2,0)`, respectively.  
We add two marriages to the return list to ensure that both print
in the correct left-right order; the 14-4 one is correct by default
but it is easier to output a line than check sex orders.  

The left panel of the figure above shows a case where
subject 11 marries into the Pedigree but also has a second spouse.
The $anchor$ variable for
this case will be `(2, 0)`; the first instance of 11 has a spouse tied
into the tree above, the second instance has no upward connections.
In the top row, subject 6 has values of `(0, 0)` since neither 
connection has an upward parent.  
In the right hand panel subject 2 has an anchor variable of `(0,1)`.


Align
==========================

The top level routine for alignment has 5 arguments

- $ped$ a Pedigree. In the case of
multiple families we loop over each family separately.
- $packed$ do we allow branches of the tree to overlap?  
If FALSE the drawing is much easier, but final drawing can
take up a huge amount of space.  
- $width$ the minimum width for a packed Pedigree. This
affects only small pedigrees, since the minimum possible
width for a Pedigree is the largest number of individiuals in
one of the generations.
- $align$ should the final step of alignment be done?  This
tries to center children under parents, to the degree possible.
- $hints$ object.  This is normally blank and auto_hint
is invoked. 

The result coming back from align is a set of vectors and
matrices:

- $n$ vector, number of entries per level
- $nid$ matrix, one row per level, numeric id of the subject plotted
here
- $pos$ the horizontal position for plotting
- $spouse$ integer matrix, one row per level, subject directly to my
right is my spouse (1), a double marriage (2), or neither (0).
- $fam$ matrix, link upward to my parents, or 0 if no link.

## Setup1 -align

Start with some setup. Throughout this routine the row number is used as a
subject id (ignoring the actual id label).

- Check that everyone has either two parents or none (a singleton confuses us).
- Verify that the hints are correct.
- The relation data frame, if present, has a factor, convert to numeric.
- Create the $spouselist$ array.  This has 4 columns
    - Husband index (4 = 4th person in the Pedigree structure)
    - Wife index
    - Plot order: 1 = husband left, 2 = wife left
    - Anchor: 1 = left member, 2 = right member, 0 = not yet determined

As the routine proceeds a spousal pair can be encountered
multiple times; we take them out of this list when the ``connected''
member is added to the Pedigree so that no marriage gets added
twice.

- To detect duplicates on the spouselist we need to create a
unique (but temporary) spouse-pair id using a simple hash.


When importing data from auto_hint, that routine's spouse matrix has column
1 = subject plotted on the left, 2 = subject plotted on the right.
The $spouselist$ array has column 1 = husband, 2 = wife. Hence the clumsy
looking ifelse below.  The auto_hint format is more congenial to users, who
might modify the output, the spouselist format easier for the code.


## Align-founders

The [align] routine does the alignment using 3 co-routines:

- $alignped1$ called with a single subject, returns the subtree
founded on this subject, as though it were the only tree
- $alignped2$ called with a set of sibs, calls alignped1 and 
alignped3 multiple times to create a joint Pedigree
- $alignped3$ given two side by side plotting structures, merge them
into a single one

Call [alignped1] sequentially with each founder pair and merge the
results.  
A founder pair is a married pair, neither of which has a father.


## Align-finish

Now finish up.
There are 4 tasks to do:

- For convenience the lower level routines kept the spouse
and nid arrays as a single object -- unpack them
- In the spouse array a 1 in position i indicates that subject
i and i+1 are joined as a marriage.  If these two have a common
ancestor change this to a 2, which indicates that a double line
should be used in the plot.
- Add twins data to the output.
- Do final alignment

### Finish align(2)
The twins array is of the same shape as the spouse and nid arrays:
one row per level giving data for the subjects plotted on that row.
In this case they are

- 0 = nothing
- 1 = the sib to my right is a monzygotic twin, 
- 2 = the sib to my right is a dizygote,
- 3 = the sib to my right is a twin, unknown zyogosity.

### Finish align(3)

At this point the Pedigree has been arranged, with the positions
in each row going from 1 to (number of subjects in the row).
(For a packed Pedigree, which is the usual case).
Having everything pushed to the left margin is not very
pretty, now we fix that.
Note that [alignped4] wants a T/F spouse matrix: it does not care
about the degree of relationship to the spouse.

alignped1
==========================

This is the first of the three co-routines.
It is called with a single subject, and returns the subtree founded
on said subject, as though it were the only tree.  
We only go down the Pedigree, not up.
Input arguments are

- $nid$ the numeric id of the subject in question
- $dad$ points to the row of the father, 0=no father in Pedigree
- $mom$ points to the row of the mother
- $level$ the plotting depth of each subject
- $horder$ orders the kids within a sibship
- $packed$ if true, everything is slid to the left
- $spouselist$ a matrix of spouses
    - col 1 = Pedigree index of the husband
    - col 2 = Pedigree index of the wife
    - col 3 = 1 : plot husband to the left, 2 = wife to the left
    - col 4 = 1 : left member is rooted here, 2 = right member, 0 = either

The return argument is a set of matrices as described in 
section [align], along with the spouselist matrix.
The latter has marriages removed as they are processed.

## alignped1 - part1

In this routine the $nid$ array consists of the final nid array + 1/2 of the
final spouse array.
The basic algorithm is simple.  

- Find all of the spouses for which $x$ is the anchor subject.  If
    there are none then return the trivial tree consisting of $x$ alone.
- For each marriage in the set, call [alignped2] on the children
    and add this to the result.

Note that the $spouselist$ matrix will only contain spouse pairs that
are not yet processed.
The logic for anchoring is slightly tricky.  First, if row 4 of
the spouselist matrix is 0, we anchor at the first opportunity, i.e. now..
Also note that if `spouselist[,3] == spouselist[,4]` it is
the husband who is the anchor (just write out the possibilities).

## alignped1 - part2

Create the set of 3 return structures, which will be matrices with
`(1 + nspouse)` columns.
If there are children then other routines will widen the result.

## alignped1 - part3

Now we have a list of spouses that should be dealt with and 
the the correponding columns of the spouselist matrix.  
Create the two complimentary lists lspouse and rspouse to denote
those plotted on the left and on the right. For someone with lots of spouses
we try to split them evenly. If the number of spouses is odd, then men should
have more on the right than on the left, women more on the right.
Any hints in the spouselist matrix override. We put the undecided marriages
closest to $x$, then add predetermined ones to the left and
right. The majority of marriages will be undetermined singletons, for which
nleft will be 1 for female (put my husband to the left) and 0 for male.
In one bug found by plotting canine data, lspouse could initially be empty but 
`length(rspouse)> 1`. This caused `nleft>length(indx)`. A fix was to not let
indx to be indexed beyond its length, fix by JPS 5/2013.

### alignped1 - part4

The spouses are in the Pedigree, now look below. For each spouse get the
list of children. If there are any we call [alignped2] to generate their tree
and then mark the connection to their parent. If multiple marriages have
children we need to join the trees.

### alignped1 - part5

To finish up we need to splice together the tree made up
from all the kids, which only has data from lev+1 down,
with the data here.  
There are 3 cases.  The first and easiest is when no
children were found.
The second, and most common, is when the tree below is
wider than the tree here, in which case we add the
data from this level onto theirs.
The third is when below is narrower, for instance an
only child.

alignped2
==========================

This routine takes a collection of siblings, grows the tree for
each, and appends them side by side into a single tree.
The input arguments are the same as those to
[alignped1] with the exception that $x$ will be a vector.
This routine does nothing to the spouselist matrix, but needs
to pass it down the tree and back since one of the routines
called by [alignped2] might change the matrix.

The code below has one non-obvious special case.  Suppose that two sibs marry.
When the first sib is processed by [alignped1] then both partners (and any
children) will be added to the rval structure below.  When the second sib is
processed they will come back as a 1 element tree (the marriage will no longer
be on the spouselist), which should **not** be added onto rval. The rule
thus is to not add any 1 element tree whose value (which must be `x[i]`) is
already in the rval structure for this level. (Where did Curt Olswold. find
these families?)

alignped3
==========================

The third alignment co-routine merges two pedigree trees which are side by side
into a single object.
The primary special case is when the rightmost person in the left
tree is the same as the leftmost person in the right tree; we  need not plot
two copies of the same person side by side. (When initializing the output
structures do not  worry about this - there is no harm if they are a column
bigger than finally needed.) Beyond that the work is simple bookkeeping.


## alignped3 - slide

For the unpacked case, which is the traditional way to draw a Pedigree when
we can assume the paper is infinitely wide, all parents are centered over their
children. In this case we think if the two trees to be merged as solid blocks.
On input they both have a left margin of 0. Compute how far over we have to
slide the right tree.

## alignped3 - merge

Now merge the two trees.  Start at the top level and work down.

- If `n2 = 0`, there is nothing to do
- Decide if there is a subject overlap, and if so 
    - Set the proper parent id. 
        Only one of the two copies will be attached and the other
        will have `fam = 0`, so `max(fam, fam2)` preserves the correct one.
    - If not packed, set the position.  Choose the one connected
        to a parent, or midway for a double marriage.
- If `packed = TRUE` determine the amount of slide for this row. It
    will be $space$ over from the last element in the left Pedigree,
    less overlap.
- Move everything over
- Fix all the children of this level, right hand Pedigree, to
    point to the correct parental position.

alignped4
==========================

The alignped4 routine is the final step of alignment.  It attempts to line
up children under parents and put spouses and siblings "close" to each other,
to the extent possible within the constraints of page width.  This routine
used to be the most intricate and complex of the set, until I realized that
the task could be cast as constrained quadradic optimization.
The current code does necessary setup and then calls the $quadprog$
function.  
At one point I investigated using one of the simpler least-squares routines
where $\beta$ is constrained to be non-negative. 
However a problem can only be translated into that form if the number
of constraints is less than the number of parameters, which is not
true in this problem.

There are two important parameters for the function.  One is the user specified
maximum width.  The smallest possible width is the maximum number of subjects
on a line, if the user suggestion is too low it is increased to that 1+ that
amount (to give just a little wiggle room). The other is a vector of 2
alignment parameters $a$ and $b$. For each set of siblings $x$ with parents
at $p_1$ and $p_2$ the alignment penalty is

$$
    (1/k^a)\sum{i=1}{k} (x_i - (p_1 + p_2)^2
$$

where $k$ is the number of siblings in the set.
Using the fact that $\sum(x_i-c)^2 = \sum(x_i-\mu)^2 + k(c-\mu)^2$,
when $a=1$ then moving a sibship with $k$ sibs one unit to the left or
right of optimal will incur the same cost as moving one with only 1 or
two sibs out of place.  If $a=0$ then large sibships are harder to move
than small ones, with the default value $a=1.5$ they are slightly easier 
to move than small ones.  The rationale for the default is as long as the
parents are somewhere between the first and last siblings the result looks
fairly good, so we are more flexible with the spacing of a large family.
By tethering all the sibs to a single spot they tend are kept close to 
each other.
The alignment penalty for spouses is $b(x_1 - x_2)^2$, which tends to keep 
them together.  The size of $b$ controls the relative importance of sib-parent
and spouse-spouse closeness.

## alignped4 - part1

We start by adding in these penalties.  The total number of parameters
in the alignment problem (what we hand to quadprog) is the set 
of `sum(n)` positions.  A work array myid keeps track of the parameter
number for each position so that it is easy to find.
There is one extra penalty added at the end.  Because the penalty amount
would be the same if all the final positions were shifted by a constant,
the penalty matrix will not be positive definite; solve.QP does not like  
this.  We add a tiny amount of leftward pull to the widest line. 

## alignped4: part2
Next come the constraints.  If there are $k$ subjects on a line there will
be $k+1$ constraints for that line.  The first point must be $\ge 0$, each
subesquent one must be at least 1 unit to the right, and the final point
must be $\le$ the max width.

Session information
===================

```{r}
sessionInfo()
```
