---
title: "Kinship2 pedigree examples"
author: "Louis Le Nézet"
date: "21/07/2023"
url: "https://github.com/LouisLeNezet/kinship2"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
header-includes: \usepackage{tabularx}
vignette: |
  %\VignetteIndexEntry{Pedigree_Examples}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

 
```{r echo = FALSE}
options(width = 100)
```

Introduction
===============

This document is a brief tutorial for the `kinship2 package`, with examples of creating pedigree objects and kinship matrices
and other pedigree utilities.
The `kinship2 package` contains the outines to handle family data with a pedigree object.
The initial purpose was to create correlation structures that describe family relationships such as kinship and identity-by-descent,
which can be used to model family data in mixed effects models, such as in the coxme function.
It also includes tools for pedigree drawing and filterinf which is focused on producing compact layouts without intervention.
Recent additions include utilities to trim the pedigree object with various criteria, and kinship for the X chromosome.

Installation
===============
The `kinship2 package` is available on Bioconductor and can be installed with the following command:

```{r}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager") }
BiocManager::install("kinship2")
```

The package can then be loaded with the following command:

```{r}
library(kinship2)
```

Table of Contents
=================

- [Introduction](#introduction)
- [Installation](#installation)
- [Table of Contents](#table-of-contents)
- [The pedigree object](#the-pedigree-object)
- [Session information](#session-information)

The pedigree object
===================

The pedigree object is a list of dataframes that describe the family structure.
It contains the following components:

- ped: a dataframe with the pedigree information
- rel: a dataframe with the relationship information
- scale: a dataframe with the affection information for the plot
- hints: a dataframe with the hints to draw the pedigree

Session information
===================

```{r}
sessionInfo()
```
