---
title: "Pedigree_Examples"
author: "Jason Sinnwell"
date: '`r format(Sys.time(),"%d %B, %Y")`'
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2
header-includes: \usepackage{tabularx}
vignette: |
  %\VignetteIndexEntry{Pedigree_Examples}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

  
```{r echo = FALSE}
options(width = 100)
```

Introduction
===============

This document is a brief tutorial for the [kinship2 package](https://www.karger.com/Article/Pdf/363105), with examples of creating pedigree objects and kinship matrices, and other pedigree 
utilities.  If the kinship2 package is not loaded, we load it now.

```{r, load, echo=TRUE}
require(kinship2)
```

Basic Usage
=============

## Example Data
Two datasets are provided within the kinship2 package: 
+ minnbreast: 17 families from a breast cancer study
+ sampleped: two samplepedigrees, with 41 and 14 subjects


This vignette uses the two pedigrees in ~sampleped~. For more information on these
datasets, see help(minnbreast) and help(sampleped).

## Pedigree

First, we load ~sampleped~ and look at some of the values 
in the dataset, and create a *pedigreeList* object using the 
*pedigree()* function.  We use the required arguments 
$id$, $father$, $mother$, and $sex$.  The $famid$ argument is 
required to make a {\em pedigreeList} object, but not for a single 
*pedigree}* object. 

```{r, pedList}
data(sampleped)
sampleped[1:10, ]
ped <- pedigree(sampleped)
print(ped)
```

The {\em pedigreeList} object can be subset to individual pedigrees 
by their family id.  The pedigree object has a print and plot method, which 
we show below.  The print method prints a short summary of the pedigree, 
while the plot in Figure~\ref{basicPed2} displays the smaller pedigree.

```{r, plot12}
ped1 <- ped[ped$ped$family == "1", ]
ped2 <- ped[ped$ped$family == "2", ]
print(ped1)
print(ped2)
plot(ped2)
plot(ped1)
```

Fixing Pedigree Issues
=======================

To "break" the pedigree, we can manipulate the sex value to not match the
parent value (in this example, we change $203$ from a male to a female, even
though $203$ is a father). To do this, we first subset ~datped2~, locate the *id* column, 
and match it to a specific id (in this case, $203$). Within id $203$, then locate in the *sex* column.
Assign this subset to the incorrect value of *2* (female) to change the original/correct
value of *1* (male).

To further break the pedigree, we can delete subjects who seem irrelevant to the
pedigree (in this example, we delete $209$ because he is a married-in father). 
To do this, we subset ~datped2~ and use the *-which()* function to locate and delete
the specified subject (in this case, $209$). Reassign this code to ~datped22~ to drop 
the specified subject entirely.

```{r, datped2}
datped2 <- sampleped[sampleped$family == 2, ]
datped2[datped2$id %in% 203, "sex"] <- 2
datped2 <- datped2[-which(datped2$id %in% 209), ]
```

An error occurs when the *pedigree()* function notices that id $203$ is not 
coded to be male (*1*) but is a father. To correct this, we simply employ the
*fix_parents()* function to adjust the *sex* value to match either *mother/momid*
or *father/dadid*. *fix_parents()* will also add back in any deleted subjects,
further fixing the pedigree.

```{r, fixped}
tryout <- try({
    ped2 <- pedigree(datped2)
})
fixped2 <- with(datped2, fix_parents(id, dadid, momid, sex))
fixped2
ped2 <- pedigree(fixped2)
plot(ped2)
```

If the fix is straightforward (changing one sex value based on either being
a mother or father), *fix_parents()* will resolve the issue. If the issue is 
more complicated, say if $203$ is coded to be both a father *and* a mother, 
*fix_parents()* will not know which one is correct and therefore the issue will 
not be resolved.

Kinship
==============

A common use for pedigrees is to make a matrix of kinship coefficients that 
can be used in mixed effect models.  A kinship coefficient is the probability 
that a randomly selected allele from two people at a given locus will be 
identical by descent (IBD), assuming all founder alleles are independent. 
For example, we each have two alleles per autosomal marker, so sampling two
alleles with replacement from our own DNA has only $p=0.50$ probability of
getting the same allele twice.  

\subsection*{Kinship for pedigree object}
We use {\em kinship} to calculate the kinship matrix for $ped2basic$. The 
result is a special symmetrix matrix class from the
[Matrix R package](https://CRAN.R-project.org/package=Matrix/),
which is stored efficiently to avoid repeating elements.

```{r, calckinship}
kin2 <- kinship(ped2basic)
kin2
```

For family 2, see that the row and column names match the id in 
the figure below, and see that each kinship coefficient with 
themselves is $0.50$, siblings are $0.25$ (e.g. $204-205$), and pedigree 
marry-ins only share alleles IBD with their children with coefficient $0.25$ 
(e.g. $203-210$). The plot can be used to verify other kinship coefficients.

## Kinship for pedigreeList object

The kinship function also works on a {\em pedigree} object with multiple families. 
We show how to create the kinship matrix, then 
show a snapshot of them for the two families, where the row and columns names
are the ids of the subject.

```{r, kin_all}
ped <- pedigree(sampleped)
kin_all <- kinship(ped)
kin_all[1:14, 1:14]
kin_all[40:43, 40:43]
kin_all[42:46, 42:46]
```

## Kinship for twins in pedigreeList object

Specifying twin relationships in a pedigree with multiple families object is complicated by the fact that the user must specify the family id to which the ~id1~ and ~id2~ belong.
We show below the relation matrix requires the family id to be in the last column, with the column names as done below, to make the plotting and kinship matrices to show up with the monozygotic twins correctly.
We show how to specify monozygosity for subjects 206 and 207 in samplepedigree 2, and subjects 125 and 126 in pedigree 1.
We check it by looking at the kinship matrix for these pairs, which are correctly at $0.5$.

```{r, kintwins}
reltwins <- as.data.frame(rbind(c(206, 207, 1, 2), c(125, 126, 1, 1)))
colnames(reltwins) <- c("indId1", "indId2", "code", "family")
ped <- pedigree(sampleped, rel_df = reltwins)

kin_all <- kinship(ped)
kin_all[21:31, 21:31]
kin_all[42:50, 42:50]
kin_all
```

Note that subject $113$ is not in pedigree 1 because they 
are a marry-in without children in the pedigree. Subject $113$ is in their own 
pedigree of size 1 in the $kin_all$ matrix at index $41$. We later show how 
to handle such marry-ins for plotting.

Optional Pedigree Informations
===============================

We use pedigree 2 from $sampleped$ to sequentially 
add optional information to the pedigree object.

## Status
The example below shows how to specify a $status$ indicator, such as 
vital status. The $sampleped$ data does not include such an  
indicator, so we create one to indicate that the first generation of 
pedigree 2, subjects 1 and 2, are deceased.  

```{r, censor}
df2 <- sampleped[sampleped$ped == 2, ]
names(df2)
df2$status <- c(1, 1, rep(0, 12))
ped2 <- pedigree(df2)
```

## Affected Indicators

We show how to specify affected status with a single indicator and 
multiple indicators in a matrix.  First, we use the affected indicator from 
$sampleped$, which contains 0/1 indicators and NA as missing, and let it
it indicate blue eyes. Next, we create a matrix to contain the 
affected indicator from $sampleped$ and a second indicator that we create, 
imagine as an indicator for baldness.

```{r, affected}
ped2 <- pedigree(df2)
aff2 <- data.frame(
  blue = df2$affected,
  bald = c(0, 0, 0, 0, 1, 0, 0, 0, 0, 1, 1, 0, 0, 1)
)
ped2 <- pedigree(df2$id, df2$father, df2$mother,
  df2$sex,
  affected = as.matrix(aff2),
  status = df2$censor
)
```

## Special Relationships

Special pedigree relationships can be specified in a matrix 
as the $relation$ argument.  There are 4 relationships that can 
be specified by numeric codes: 1=Monozygotic twins, 2=Dizygotic twins, 
3=twins of unknown zygosity, and 4=Spouse. The spouse relationship can 
indicate a marry-in when a couple does not have children together.

Below, we create a matrix of relationships for monozygotic and 
unknown-zygosity twins in the most recent generation of pedigree 2.  

```{r, twins}
## create twin relationships
relate2 <- matrix(c(
  210, 211, 1,
  212, 213, 3
), nrow = 2, byrow = TRUE)
ped2 <- pedigree(df2$id, df2$father, df2$mother,
  df2$sex,
  affected = as.matrix(aff2),
  status = df2$censor,
  relation = relate2
)
```

Another special relationship is inbreeding. Inbreeding of founders implies the founders' parents are related (the maternal and paternal genes descended from a single ancestral gene). One thing we can do is add more people to the pedigree to show this inbreeding.

To show that a pair of founders (subjects $201$ and $202$) are inbred, we must show that their parents are siblings. To do this, we create subjects $197$ and $198$ to be the parents of $201$ and also create subjects $199$ and $200$ to be the parents of $202$. To make subjects $198$ and $199$ siblings, we give *them* the same parents, creating subjects $195$ and $196$. This results in subjects $201$ and $202$ being first cousins, and therefore inbred.

```{r, inbreeding}
id <- 195:202
dadid <- c(0, 0, 0, 196, 196, 0, 197, 199)
momid <- c(0, 0, 0, 195, 195, 0, 198, 200)
sex <- c(2, 1, 1, 2, 1, 2, 1, 2)
ped3 <- pedigree(id, dadid, momid, sex)
ped4df <- rbind.data.frame(as.data.frame(ped2)[-c(1, 2), 1:4], as.data.frame(ped3))
ped4 <- with(ped4df, pedigree(id, dadid, momid, sex))
plot(ped4, cex = .5)
```

Pedigree Plot Details
===========================

The plot method does an admirable job plotting pedigrees within the
standard R plotting paradigm.  It attempts to adhere to many standards
in pedigree plotting, as presented by [Bennet et al. 2008](https://pubmed.ncbi.nlm.nih.gov/18792771/).

We show in the following figure, the plot of the updated pedigree 2.
The plot shapes for each subject are divided into two equal parts and
shaded differently to indicate the two affected indicators.
Also, the two deceased subjects are displayed with a diagonal line 
through the shape. The twin relationships are both represented with 
diverging lines from a single point.  The monozygotic twins have an 
additional line connecting the diverging lines, while the other twins have 
a question mark to indicate unknown zygosity.

We also show how the subjects can be colored individually, where we 
color a subject red if their ~avail~ indicator is 1, which can 
represent their DNA availability, a useful indicator in genetic studies.
Lastly, we show how to use the $id$ argument in the plot method to add 
additional information under each subject. In the example below, we add 
names to the existing $id$ vector using the newline character as the 
$sep$ argument in *paste()*.  As space permits, more lines 
and characters per line can be made using the $id$ argument.

Finally, we show how a pedigree.legend can place a simple legend in one
of the corners of the pedigree plot to show the sections of the plot 
symbols corresponding to the multiple affected indicators.

```{r, ped2update}
id2 <- paste(df2$id, c(
  "John", "Linda", "Jack", "Rachel", "Joe", "Deb",
  "Lucy", "Ken", "Barb", "Mike", "Matt",
  "Mindy", "Mark", "George"
), sep = "\n")
plot(ped2, avail = ifelse(df2$avail, 2, 1), id = id2, cex = .5)
#pedigree.legend(ped2, location = "topright", radius = .2)
```

To show some other tricks with pedigree plotting, we use pedigree 1 
from ~sampleped~, which has 41 subjects in 4 generations, including a
generation with double first cousins. After the first marriage of 114, they
remarried subject 113 without children between them.  If we do not 
specify the marriage with the $relation$ argument, the plot method excludes
subject $113$ from the plot. The basic plot of pedigree 1 is shown 
in the figure below, where the subjects are colored red if their 
~avail~ indicator is 1.


```{r, plotped1}
df1 <- sampleped[sampleped$ped == 1, ]
relate1 <- matrix(c(113, 114, 4), nrow = 1)
ped1 <- pedigree(df1$id, df1$father, df1$mother,
  df1$sex,
  affected = df1$affected,
  relation = relate1
)
print(ped1)
plot(ped1, avail = ifelse(df1$avail == 1, "red", "black"), cex = .7)
```

## Align by Input Order

The plot method does a decent job aligning subjects given the order of the 
subjects when the pedigree object is made, and sometimes has to make two 
copies of a subject.  If we change the order of the subjects when creating 
the pedigree, we can help the plot method reduce the need to duplicate 
subjects, as Figure~\ref{reordPed1} no longer has subject $110$ duplicated.

```{r, ordering}
df1reord <- df1[c(35:41, 1:34), ]
ped1reord <- pedigree(df1reord$id, df1reord$father, df1reord$mother,
  df1reord$sex,
  affected = df1reord$affected, relation = relate1
)
plot(ped1reord, avail = df1reord$avail + 1, cex = .7)
```
  
## Legend Plot

The *legendPlot()* function creates a dynamic legend based on the arguments
passed through it. First, we give it color names and new labels for the affected
status, which is two for blue eyes and baldness that we created previously for ped2.

```{r, legendplot, eval = FALSE}
legendPlot(ped2,
  col = ifelse(df2$avail, "red", "black"),
  id = id2, symbolsize = .4,
  affected.label = c("blue eyes", "baldness"),
  col.label = c(`black` = "no dna", `red` = "dna")
)
```

We can do the same plot by giving integer color codes, and a non-named vector of
colors. Also, we can let the function grab the colnames() of the affected matrix
for the legend.

```{r, legendplotindex, eval = FALSE}
colnames(ped2$affected)
legendPlot(ped2,
  col = ifelse(df2$avail, 2, 1),
  id = id2, cex = .5, symbolsize = .5,
  col.label = c("no dna", "dna")
)
```
  
Pedigree Utility Functions
=================================

## Pedigree as a Data.Frame

A main features of a pedigree object are vectors with an element
for each subject. It is sometimes useful to extract these vectors from 
the pedigree object into a $data.frame$ with basic information that can
be used to construct a new pedigree object. This is possible with the
$as.data.frame()$ method, as shown below.

```{r, ped2df}
dfped2 <- as.data.frame(ped2)
dfped2
```

## Subsetting and Trimming

Pedigrees with large size can be a bottleneck for programs that run 
calculations on them. The kinship2 package contains some routines to 
identify which subjects to remove.  We show how a subject 
(e.g. subject 210) can be removed from 
~ped2~, and how the pedigree object is changed by verifying that the
~relation~ matrix no longer has the twin relationship between subjects 
210 and 211, as indicated by $indx1$ and ~indx2~. Also note that
the $relation$ matrix indices are updated for persons 212 and 213 who
have index 11 and 12 after subject 210 is removed.

```{r, subset}
ped2.rm210 <- ped2[-10]
data.frame(ped2.rm210)
ped2.rm210$relation
ped2$relation
```

The steps above only work for subsetting by the index of the 
pedigree object vectors, not by the ~id~ of the subjects themselves.  
We provide *pedigree_trim()*, which trims subjects from a pedigree by 
their $id$.  Below is an example of removing subject 110, as done above, 
then we further trim the pedigree by a vector of subject ids. We check the
trimming by looking at the $id$ vector and the $relation$ matrix.

```{r, trim}
ped2.trim210 <- pedigree_trim(210, ped2)
ped2.trim210$id
ped2.trim210$relation
ped2.trim.more <- pedigree_trim(c(212, 214), ped2.trim210)
ped2.trim.more$id
ped2.trim.more$relation
```


Shrinking
==============

An additional function in kinship2 is *pedigree.shrink()*, which shrinks a pedigree
to a specified bit size while maintaining the maximal amount of 
information for genetic linkage and association studies.  Using an indicator 
for availability and affected status, it removes subjects in this order:
  + unavailable with no available descendants
  + available and are not parents
  + available who have missing affected status
  + available who are unaffected
  + available who are affected

We show how to shrink pedigree 1 to bit size $30$, which happens 
to be the bit size after removing only the unavailable subjects.  We show how 
to extract the shrunken pedigree object from the $pedigree.shrink$ result, 
and plot it.

```{r, shrink1, eval=FALSE}
set.seed(200)
shrink1.B30 <- pedigree.shrink(
  ped = ped1,
  avail = df1$avail, maxBits = 30
)
print(shrink1.B30)
# plot.pedigree.shrink(shrink1.B30, avail=shrink1.B30$avail + 1, cex=.6)
```

Now shrink pedigree 1 to bit size $25$, which requires removing 
subjects who are informative.  If there is a tie between multiple subjects 
about who to remove, the method randomly chooses one of them. With this 
seed setting, the method removes subjects $126$ then $125$.

```{r, shrink2, eval=FALSE}
set.seed(10)
shrink1.B25 <- pedigree.shrink(
  ped = ped1, avail = df1$avail,
  maxBits = 25
)
shrink1.B25
# plot.pedigree.shrink(shrink1.B25, avail=shrink1.B25$avail+1, cex=.6)
```

Select Unrelateds
=======================

In this section we briefly show how to use $unrelated$ to find
a set of the maximum number of unrelated available subjects from a pedigree.
The input required is a pedigree object and a vector indicating availability.
In some pedigrees there are numerous sets of subjects that satisfy the maximum
number of unrelateds, so the method randomly chooses from the set. We 
show two sets of subject ids that are selected by the routine and discuss 
below.

```{r, unrelateds}
df2 <- sampleped[sampleped$ped == 2, ]
ped2 <- pedigree(df2$id, df2$father, df2$mother,
  df2$sex,
  affected = df2$affected
)
set.seed(10)
set1 <- unrelated(ped2, avail = df2$avail)
set1
set2 <- unrelated(ped2, avail = df2$avail)
set2
```

We can easily verify the sets selected by $unrelated$ by
referring to Figure~\ref{basicPed2} and see that subjects 203 and 209 are
unrelated to everyone else in the pedigree except their children. Furthermore, 
we see in df2 that of these two, only subject 203 is available. Therefore,
any set of unrelateds who are available must include subject
203 and one of the these subjects: 201, 204, 206, 207, 212, 
and 214, as indicated by the kinship matrix for pedigree 2 subset to those 
with availability status of 1.

```{r, unrelVerify}
df2
kin2[df2$avail == 1, df2$avail == 1]
```
